---
title: "PA1_template1"
author: "pedro pablo malagon amor"
date: "Saturday, July 19, 2014"
output: html_document
---
# Reproducible Research: Peer Assessment 1
first assignment for the Reproducible Research course on Coursera. 
objective  write a report using R markdown that can be processed by knitr package and be transformed into an HTML file. 


## Loading and preprocessing the data
set the work directory `setwd()`. `raw_data` contains a data frame with the raw data used 

data consists of  two data sets: 
one for **total number of steps per day** and 
one for **average of steps per time interval**. 
Both data sets were generated using the `aggregate` function and were assigned to the objects `avg.daily_data` and `avg.interval_data`, respectively.

```{r}
## Set work directory
path_mydir = "C:\\Users\\pedropm\\Documents\\prog1rs\\"
setwd(path_mydir)

## Load data
raw_data = read.csv("activity.csv", header=TRUE)

## Dataset: total steps per day
avg.daily_data        = aggregate(raw_data$steps, by=list(raw_data$date), sum, na.rm=TRUE)
# Column names
names(avg.daily_data) = c("date", "total.steps")

## Dataset: average steps per interval
# Interval as factor
raw_data$interval        = as.factor(raw_data$interval)
# Dataset
avg.interval_data        = aggregate(raw_data$steps, by=list(raw_data$interval), mean, na.rm=TRUE)
# Column names
names(avg.interval_data) = c("interval","avg.steps")
```


## What is mean total number of steps taken per day?

The `avg.daily_data` data set  used to answer. 
The mean total number of steps taken per day was assigned to the object `total.steps_mean` and the median was assigned to the `total.steps_median`.

```{r, fig.width=12}
## Histogram of the total number of steps taken each day
hist(avg.daily_data$total.steps, xlab="Total number of steps taken per day", breaks=20, col="lightblue", border="darkgrey", main="")

## Mean and median of the total number of steps taken per day
total.steps_mean   = mean(avg.daily_data$total.steps)
total.steps_mean
total.steps_median = median(avg.daily_data$total.steps)
total.steps_median
```

The mean total number of steps taken per day is equal to **`r round(total.steps_mean)`**, whereas the median is **`r total.steps_median`**. 


## What is the average daily activity pattern?

The `avg.interval_data` data set was used to answer this question. 


```{r, fig.width=21}
## Time series plot of the 5-minute interval and the average number of steps taken
# Graph margins
par(mar=c(6,6,1.5,1.5))
plot(1:288, avg.interval_data$avg.steps, xlab="Time interval", ylab="Average steps per time interval", xlim=c(0,288), ylim=c(0,max(avg.interval_data$avg.steps)+20), xaxs="i", xaxt="n", type="l", col="deepskyblue", cex=1.5, cex.lab=1.5)
# Labels
atlabel = seq(1, 287, by=2)
clabel  = levels(avg.interval_data$interval)
clabel  = clabel[atlabel]
axis(1, at=atlabel, labels=clabel, cex.lab=0.2)
# Plot line at interval with max avg. step value
abline(v=which.max(avg.interval_data$avg.steps), col="grey", lty=2)

## Time interval with largest average of steps
interval_max.steps = avg.interval_data[which.max(avg.interval_data$avg.steps),]
interval_max.steps
```

On average across all the days in the dataset, the 5-minute interval **`r interval_max.steps$interval`** contains the maximum number of steps. 

## Imputing missing values

The missing values were totalized based on the number of rows from the `raw_data` data set which contained *NA* values in the *steps* column. The *NA* values were firstly filled using the `na.approx()` function from the `{zoo}` package. 
A new data set was created with *NA* filled, named `raw_data_filled`, which was used to calculate the average total steps per day that was assigned to the object `avg.daily_data_filled`.

```{r, fig.width=12}
## Total number of missing values in the raw dataset
nrow(raw_data[is.na(raw_data$steps)==TRUE,])

## Filling missing values in the raw dataset
# install.packages("zoo") # if you don't have it installed yet
library(zoo)
raw_data_filled       = raw_data
# Fill dataset using interpolation
raw_data_filled$steps = na.approx(raw_data_filled$steps, na.rm=FALSE)
# Intervals that are still missing
still_na              = raw_data_filled[is.na(raw_data_filled$steps)==TRUE,]
still_na              = match(still_na$interval, avg.interval_data$interval)
still_na              = avg.interval_data$avg.steps[still_na]
# Fill still missing values with average of that interval
raw_data_filled$steps[is.na(raw_data_filled$steps)==TRUE] = still_na

## Dataset: total steps per day
avg.daily_data_filled = aggregate(raw_data_filled$steps, by=list(raw_data_filled$date), sum, na.rm=TRUE)
names(avg.daily_data_filled) = c("date", "total.steps")

## Histogram of the total number of steps taken each day with filled NA dataset
hist(avg.daily_data_filled$total.steps, xlab="Total number of steps taken per day", breaks=20, col="lightblue", border="darkgrey", main="")

## Mean and median of the total number of steps taken per day with filled NA dataset
total.steps_mean_filled   = mean(avg.daily_data_filled$total.steps)
total.steps_mean_filled
total.steps_median_filled = median(avg.daily_data_filled$total.steps)
total.steps_median_filled
```

The missing values totals **`r nrow(raw_data[is.na(raw_data$steps)==TRUE,])`**. The mean total number of steps taken per day with filled gaps is equal to **`r round(total.steps_mean_filled)`**, whereas the median is **`r total.steps_median_filled`**. 

## Are there differences in activity patterns between weekdays and weekends?

Higher values of the average steps per time are concentrated during weekdays, when the highest average number of steps per time interval occurs. During weekends, the average steps per time are more even distribuited across time intervals.

```{r, fig.width=21}
## Add Type variable to raw data filled that contains wheather date is weekday or weekend
raw_data_filled$Type = weekdays(as.Date(raw_data_filled$date))
raw_data_filled$Type = ifelse(raw_data_filled$Type == "Saturday" | raw_data_filled$Type == "Sunday", "weekend", "weekday")
raw_data_filled$Type = as.factor(raw_data_filled$Type)

## Dataset: average steps per interval with filled data
raw_data_filled$interval        = as.factor(raw_data_filled$interval)
avg.interval_data_filled        = aggregate(raw_data_filled$steps, by=list(raw_data_filled$interval, raw_data_filled$Type), mean, na.rm=TRUE)
names(avg.interval_data_filled) = c("interval", "Type", "avg.steps")

## Time series panel plot of 5-minute interval (x-axis) and the average number of steps taken give weekday type
# Labels
atlabel = seq(1, 287, by=6)
clabel  = levels(avg.interval_data$interval)
clabel  = clabel[atlabel]
# Plot
library(lattice)
xyplot(avg.steps ~ interval | Type, data=avg.interval_data_filled, layout=c(1,2), type="l", ylab="Average steps per time interval", col="deepskyblue", scales=list(x=list(at=atlabel, labels=clabel)))
```
